---
title: "Tidying Public Water Quality Data"
author: "Matthew Ross"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
editor_options: 
  markdown: 
    wrap: 72
---

# Why public data sets?

Working with large, open-access data sets can serve many purposes. It
can be an excellent way to explore new ideas, before investing in
field-work or experiments. It can be a great way to take local or
experimental results and expand them to different ecosystems, places, or
landscapes. Or, it can be an excellent way to build, validate, and test
ecological models on regional or national scales.

So why doesn't everyone use public data? Well, it's often collected by a
variety of organizations, with different methods, units, and
inconsistent metadata. Together these issues make large public data sets
quite messy. This messiness can take many different forms, but at the
basic level it means that data is hard to analyze. This is not
necessarily because the data itself is bad, but because the way it is
organized is unclear and/or inconsistent.

In this assignment, we will use the skills we've learned to convert a
messy data set into something ready for analysis. As always, we will
depend heavily on the {tidyverse}, an excellent series of packages that
make data manipulation beautiful and easy. We will also be working with
the US's [Water Quality Portal data
base](https://www.waterqualitydata.us/), which we will access with the
{dataRetrieval} package.

## Loading key packages

Let's first load in all of our necessary packages, which you can do with
the 'setup.R' script we've created for you. Some of these packages might
be new to you, such as:

-   `mapview` : Similar to `tmap`, another package that allows us to
    interactively map our data

-   `broom` : simplified model outputs

-   `trend` : Package to explore trends in data

```{r setup, warnings = 'hide', message = FALSE}
source('setup.R')
```

# Downloading data

For this lesson, we'll explore water quality data in the Colorado River
Basin as it moves from Colorado to Arizona. All data will be generated
through the code you see below, with the only external information
coming from knowing the site IDs for the monitoring locations along the
Colorado River and the water quality characteristic names we are
interested in.

The water quality portal can be accessed with the function
`readWQPdata()`, which takes a variety of arguments (like start date,
end date, constituents, etc.). We'll generate these arguments for
downloading the data below.

## Download prep

First we'll make a tibble (aka, a tidyverse-style table/data frame) of
our site IDs of interest. (Generally, as the site ID's number increases,
the site's location moves further downstream of the river's headwaters,
which are near Grand Lake, CO.)

```{r}
colorado <- tibble(siteid = c("USGS-09034500", "USGS-09069000",
                              "USGS-09085000", "USGS-09095500", "USGS-09152500"),
                   basin = c("colorado1", "eagle",
                             "roaring", "colorado3", "gunnison"))
```

Now that we have a tibble that contains a list of our sites of interest,
we next will need to set up a series of rules for downloading data from
the Water Quality Portal.

We'll focus on cation and anion data from 1980-present. Each cation has
a name that we might typically use (like calcium or sulfate), but the
name may be different in the Water Quality Portal, so we have to check
this website
(<https://www.waterqualitydata.us/Codes/Characteristicname?mimeType=xml>)
to get our names correct.

```{r}
# ca = "Calcium"
# mg = "Magnesium"
# na = "Sodium"
# k = "Potassium"
# so4 = c("Sulfate", "Sulfate as SO4", "Sulfur Sulfate", "Total Sulfate")
# cl = "Chloride"
# hco3 = c("Alkalinity, bicarbonate", "Bicarbonate")

# Compile all these names into a single vector:
parameters <- c("Calcium", "Magnesium", "Sodium", "Potassium", "Sulfate", "Sulfate as SO4", "Sulfur Sulfate", "Total Sulfate", "Chloride", "Alkalinity, bicarbonate", "Bicarbonate")
```

Using our site names and characteristic names generated above, we can
now run the `readWQPdata()` function to import water quality data for
each of our sites and parameters:

```{r}
conc_wide <- readWQPdata(siteid = colorado$siteid, # pulling our siteids column in from the colorado object, becomes a vector
                         sampleMedia = "Water", # WQP also has bilogical data and sediment data
                         startDateLo = "1980-10-01", # must be formatted as YYYY-MM-DD
                         startDateHi = "2023-11-01", # must be formatted as YYYY-MM-DD
                         characteristicName = parameters) # our vector of `characteristcName`s
```

# Data tidying

Now that we have downloaded the data, we need to tidy it up. The Water
Quality Portal data comes with an incredible amount of metadata in the
form of extra columns. But we don't need all this extra data.

## Look at the data you downloaded

```{r}
View(conc_wide)
```

## Initial cleaning up

Wow, that looks messy! Lots of extraneous columns, lots of NAs, so much
information we can hardly parse it. Let's pare it down to the
essentials.

```{r}
# This code mostly just grabs and renames the most important data columns
conc_small <-  conc_wide %>%
  select(date = ActivityStartDate,
         parameter = CharacteristicName,
         units = ResultMeasure.MeasureUnitCode,
         siteid = MonitoringLocationIdentifier,
         org = OrganizationFormalName,
         org_id = OrganizationIdentifier,
         time = ActivityStartTime.Time,
         value = ResultMeasureValue,
         sample_method = SampleCollectionMethod.MethodName,
         analytical_method = ResultAnalyticalMethod.MethodName,
         particle_size = ResultParticleSizeBasisText,
         date_time = ActivityStartDateTime,
         media = ActivityMediaName,
         sample_depth = ActivityDepthHeightMeasure.MeasureValue,
         sample_depth_unit = ActivityDepthHeightMeasure.MeasureUnitCode,
         fraction = ResultSampleFractionText,
         status = ResultStatusIdentifier) %>%
  # Remove trailing white space in labels
  mutate(units = trimws(units))  
```

Let's also add in the aliases we established in our`colorado` object,
and then add a new column, `ion`, that represents what each of our
`parameter`s are associated with. Next, let's make sure this information
lives at the beginning of the table:

```{r}
conc_meta <- conc_small %>%
  left_join(., colorado, by = "siteid") %>%
  dplyr::mutate(ion = dplyr::case_when(parameter == "Calcium" ~ "Ca",
                                       parameter == "Magnesium" ~ "Mg",
                                       parameter == "Sodium" ~ "Na",
                                       parameter == "Potassium" ~ "K",
                                       parameter %in% c("Sulfate", "Sulfate as SO4", "Sulfur Sulfate", "Total Sulfate") ~ "SO4",
                                       parameter == "Chloride" ~ "Cl",
                                       parameter %in% c("Alkalinity, bicarbonate", "Bicarbonate") ~ "HCO3")) %>%
  select(siteid, basin, ion, parameter, date, everything())
```

Now let's look at this tidier version:

```{r}
View(conc_meta)
```

## Final tidy data set

That is getting better, but we still have lots of extraneous
information. For our purposes let's assume that the sample and analytic
methods used by the USGS are reasonable and exchangeable (i.e., one
method is equivalent to the other). If we make that assumption then the
only remaining data we need to clean is to make sure that all the data
has the same units.

### Unit check

```{r unit check}
table(conc_meta$units)
```

Wow! Almost all the data is in mg/L (or, depending on when you pulled
it, all of it). That makes our job really easy.

We just need to remove these (potential) non-mg/L observations with a
`dplyr::filter` call and then select an even smaller subset of useful
columns, while adding a date object column using the `lubridate::ymd`
call.

```{r tidy}
conc_tidy <- conc_meta %>% 
  filter(units == 'mg/l') %>%
  mutate(date = ymd(date)) %>%
  select(date,
         parameter,
         ion,
         siteid,
         basin,
         conc = value)
```

### Daily data

We now have a manageable data frame. But how do we want to organize the
data? Since we are looking at a really long time-series of data, let's
look at data as a daily average. The `dplyr::group_by`
`dplyr::summarize` functions make this really easy:

```{r}
# The amazing group_by function groups all the data so that the summary
# only applies to each subgroup (siteid, date, and parameter combination).
# So in the end you get a daily average concentration for each siteid and parameter type. 
conc_daily <- conc_tidy %>%
  group_by(date, parameter, siteid, basin) %>% 
  summarize(conc = mean(conc, na.rm = T))
```

Taking daily averages looks like it did eliminate
`r nrow(conc_tidy) - nrow(conc_daily)` observations, meaning these
site-date combinations had multiple observations on the same day.

# Assignment!

Let's imagine you wanted to add more data to your water quality
analyses, but you also know that you need to do this analysis over and
over again. Let's walk through how we would: 1) Add new data to our
current `conc_tidy` data set and 2) how to write a function to download,
clean, and update our data with far less code.

## Question 1

Write a function that can repeat the above steps for any table of site
IDs with a single function call. This function should take in a single
tibble that is identical in structure to the `colorado` one above (i.e.,
it has columns named `siteid` and `basin`), and produce an identical
data frame to `conc_daily` (i.e., a data frame of daily average ion
concentrations).

```{r}
download_clean_data <- function(site_data) {
  parameters <- c("Calcium", "Magnesium", "Sodium", "Potassium", "Sulfate", "Sulfate as SO4", "Sulfur Sulfate", "Total Sulfate", "Chloride", "Alkalinity, bicarbonate", "Bicarbonate")

  conc_wide <- readWQPdata(siteid = site_data$siteid,
                           sampleMedia = "Water",
                           startDateLo = "1980-10-01",
                           startDateHi = "2023-11-01",
                           characteristicName = parameters)

  conc_small <- conc_wide %>%
    select(date = ActivityStartDate,
           parameter = CharacteristicName,
           units = ResultMeasure.MeasureUnitCode,
           siteid = MonitoringLocationIdentifier,
           org = OrganizationFormalName,
           org_id = OrganizationIdentifier,
           time = ActivityStartTime.Time,
           value = ResultMeasureValue,
           sample_method = SampleCollectionMethod.MethodName,
           analytical_method = ResultAnalyticalMethod.MethodName,
           particle_size = ResultParticleSizeBasisText,
           date_time = ActivityStartDateTime,
           media = ActivityMediaName,
           sample_depth = ActivityDepthHeightMeasure.MeasureValue,
           sample_depth_unit = ActivityDepthHeightMeasure.MeasureUnitCode,
           fraction = ResultSampleFractionText,
           status = ResultStatusIdentifier) %>%
    mutate(units = trimws(units))

  conc_meta <- conc_small %>%
    left_join(., site_data, by = "siteid") %>%
    mutate(ion = case_when(parameter == "Calcium" ~ "Ca",
                           parameter == "Magnesium" ~ "Mg",
                           parameter == "Sodium" ~ "Na",
                           parameter == "Potassium" ~ "K",
                           parameter %in% c("Sulfate", "Sulfate as SO4", "Sulfur Sulfate", "Total Sulfate") ~ "SO4",
                           parameter == "Chloride" ~ "Cl",
                           parameter %in% c("Alkalinity, bicarbonate", "Bicarbonate") ~ "HCO3")) %>%
    select(siteid, basin, ion, parameter, date, everything())

  conc_tidy <- conc_meta %>%
    filter(units == 'mg/l') %>%
    mutate(date = ymd(date)) %>%
    select(date, parameter, ion, siteid, basin, conc = value)

  conc_daily <- conc_tidy %>%
    group_by(date, parameter, siteid, basin) %>% 
    summarize(conc = mean(conc, na.rm = TRUE))

  return(conc_daily)
}

```

## Question 2

Using the function you developed in Question 1, download and clean water
quality data for the site IDs listed below:

```{r}
additional_data <- tibble(siteid = c('USGS-09180000', 'USGS-09180500', 'USGS-09380000'),
                          basin = c('dolores', 'colorado4', 'colorado5'))
additional_data_result <- download_clean_data(additional_data)
```

This output of running the function should look identical in format to
our original `conc_daily` data set, but for these sites instead.

## Question 3

Combine the data pulled in Question 2 with the original data from
`conc_daily`, so that this data is in a single data frame. Save this
combined data as `tidied_full_wq.RDS` in a folder called data.

```{r}

wq <- bind_rows(conc_daily, additional_data_result)
saveRDS(wq, file = 'data/tidied_full_wq.RDS')
```

## Question 4

We now have a data set of stream water quality data for several sites
throughout Colorado. One potential control on stream chemistry is stream
discharge. A function in the {dataRetrieval} package that allows you to
easily download discharge data is `readNWISdv()`. Use this function to
download daily discharge data for all eight of the sites we are
interested in. Save the data as an RDS object called called `Q` in the
data folder: `data/Q.RDS`. The site numbers are the same as above but
you need to remove `USGS-` from each site. Discharge is
`parameterCd = 00060` and you should use `renameNWISColumns()` to
automatically make the column names a little less annoying.

```{r}
sites <- colorado %>%
  bind_rows(additional_data) %>%
  pull(siteid) %>%
  gsub('USGS-', '', .)

q_data <- readNWISdv(siteNumbers = sites,
                     parameterCd = '00060',
                     startDate = '1980-10-01',
                     endDate = '2023-11-01') %>%
  renameNWISColumns()

saveRDS(q_data, file = 'data/Q.RDS')

```
